Reproducing "Fluctuation Domains

### 2020-02-19 rescience/article.Rmd
rebuild with spellcheck :abc:

---
title: 'Reproducing "Fluctuation Domains"'
output:
  pdf_document:
    latex_engine: xelatex
    template: rescience.latex
    includes:
      before_body: header.tex
      after_body: suffix.tex
    keep_tex: true
documentclass: rescience
bibliography: "bibliography.bib"

@@

In this analsyis I seek to reproduce the numerical results of [@Boettiger2010], which focuses on the evolutionary trajectory of species trait under selection, as predicted by the so-called "Canonical Equation" of adaptive dynamics [@Dieckmann1996].  Theoretical work I present in the paper predicts the emergence of a 'flutation enhancement' regime far from an evolutionary stable point, and suggests that the evolution of traits which start in this distant region would deviate substantially from the expected evolutionary path predicted by the canonical equation.  In particular, the distribution of such trajectories could be bimodal, making the "canonical" path far from the most likely path.  The details of this distribution cannot be solved for analytically however, so in this paper I resorted to an replicate exact stochastic simulations of the evolutionary model using a Gillespie algorithm [@Gillespie].

@@

Notably, these versions are not entirely the same.  The GitHub/Zenodo/Dryad version has the addition of a `configure.ac` script for autoconf, intended to make the package more portable in assisting with the linkage to the GSL library, which is missing from the original deposit (on the GeorgiaTech website and Dryad).  The GitHub version also lacked a copy of the `inst/` subdirectory, which contained a copy of the manuscript `.tex` file, original manuscript `.pdf` figures, and the R scripts that were supposed to generate these figures. (Though both versions include a `demo` directory that included similar code required to generate the figures, as discussed in the next section.)

@@

The second challenge was in figuring out just what commands had to be run to duplicate the figures.  As mentioned above, the GitHub version of the source-code did not have the manuscript source, and even the version that did have the manuscript source did not write the manuscript itself as a dynamic document (e.g. in Sweave [@knitr_book]).  In that version, files `fig1.R` and `fig2.R` are were included in the same subdirectory as the `pdf` figures referenced by the `tex` version of the manuscript, but these files contained only the code required to actually generate the figures themselves once the result data had been pre-computed and saved as `.txt` files. The result data files had also not been archived separately.  Fortunately, re-reading my own documentation, I was able to deduce that these text files were generated by running the function `logistic()` with the parameter settings given in the caption of Fig 2 in the manuscript.

@@

Figure 2 of the original paper has the core computational results demonstrating the consequences of these fluctuation regimes.  The code for this analysis was implemented as pure C code, which can be run in stand-alone mode and generates three output text files.  The R package uses the `.C` interface to pass parameters to the C function (`Rlogistic`) which runs the ensemble and writes its output as text. The C function does not expose a random seed, so re-running the analysis creates results which are not bitwise identical each time. Of course the scientific results are expected to be independent of the seed anyway, so it is perhaps reassuring that as long as the ensemble of replicates is large enough, the same pattern can be reproduced quite closely. The original manuscript uses ensembles of 100,000 replicates, which I have reproduced here.  A possible source of confusion is that the code defaults to only 100 replicates if the ensemble size is not explicitly verified -- which is an ensemble small enough to still see random deviations from the predicted results.   I probably set that lower default to facilitate faster testing.  Likewise, the MAXTIME parameter has to be extended from the default setting to allow the third initial condition set to run sufficiently long.

@@

Note that these tighter integrations would also prevent the C libraries from running as purely stand-alone implementation.  In any event, meeting all the automated checks of a formal package is not an expectation of the compendium approach, which merely borrows some of the package tooling.

@@

In conclusion, I have been able to successfully replicate the original results.  As the original results were not performed with a fixed random seed, replicated results are statistically comparable but not bitwise identical.  In a scientific sense this is perhaps more useful than replication with a fixed seed.  My compendium approach did not install entirely out of the box, as an autoconf configure script first required updating.  Ironically, the original version of the code published on a co-author's personal website and linked from the manuscript both still existed and did install out-of-the-box.  Reproducing the results of the paper required some careful reading of parameters listed in the paper and manually knowing what functions to call, as this paper did not include a complete "dynamic document" [@knitr; @knitr_book; @rmarkdown] approach discussed in @Marwick and also used in this replication paper.  I think this experience underscored the critical role dependency management has in frustrating reproducibility: this replication was relatively simple in part due to the low number of dependencies and the technical simplicity of the code.  The other take-away for me is how easy it is to leave out small but important steps needed to reproduce a figure when not using the dynamic document approach.  Despite having R scripts called "figure1.R" and "figure2.R", these scripts could not simply be run by themselves to create the figures.

>>>>

---
title: 'Reproducing "Fluctuation Domains"'
output:
  pdf_document:
    latex_engine: xelatex
    template: rescience.latex
    includes:
      before_body: header.tex
      after_body: suffix.tex
documentclass: rescience
bibliography: "bibliography.bib"

@@

In this analysis I seek to reproduce the numerical results of [@Boettiger2010], which focuses on the evolutionary trajectory of species trait under selection, as predicted by the so-called "Canonical Equation" of adaptive dynamics [@Dieckmann1996].  Theoretical work I present in the paper predicts the emergence of a 'fluctuation enhancement' regime far from an evolutionary stable point, and suggests that the evolution of traits which start in this distant region would deviate substantially from the expected evolutionary path predicted by the canonical equation.  In particular, the distribution of such trajectories could be bimodal, making the "canonical" path far from the most likely path.  The details of this distribution cannot be solved for analytically however, so in this paper I resorted to an replicate exact stochastic simulations of the evolutionary model using a Gillespie algorithm [@Gillespie].

@@

Notably, these versions are not entirely the same.  The GitHub/Zenodo/Dryad version has the addition of a `configure.ac` script for autoconf, intended to make the package more portable in assisting with the linkage to the GSL library, which is missing from the original deposit (on the GeorgiaTech website and Dryad).  The GitHub version also lacked a copy of the `inst/` sub-directory, which contained a copy of the manuscript `.tex` file, original manuscript `.pdf` figures, and the R scripts that were supposed to generate these figures. (Though both versions include a `demo` directory that included similar code required to generate the figures, as discussed in the next section.)

@@

The second challenge was in figuring out just what commands had to be run to duplicate the figures.  As mentioned above, the GitHub version of the source-code did not have the manuscript source, and even the version that did have the manuscript source did not write the manuscript itself as a dynamic document (e.g. in Sweave [@knitr_book]).  In that version, files `fig1.R` and `fig2.R` are were included in the same sub-directory as the `pdf` figures referenced by the `tex` version of the manuscript, but these files contained only the code required to actually generate the figures themselves once the result data had been pre-computed and saved as `.txt` files. The result data files had also not been archived separately.  Fortunately, re-reading my own documentation, I was able to deduce that these text files were generated by running the function `logistic()` with the parameter settings given in the caption of Fig 2 in the manuscript.

@@

Figure 2 of the original paper has the core computational results demonstrating the consequences of these fluctuation regimes.  The code for this analysis was implemented as pure C code, which can be run in stand-alone mode and generates three output text files.  The R package uses the `.C` interface to pass parameters to the C function (`Rlogistic`) which runs the ensemble and writes its output as text. The C function does not expose a random seed, so re-running the analysis creates results which are not bit-wise identical each time. Of course the scientific results are expected to be independent of the seed anyway, so it is perhaps reassuring that as long as the ensemble of replicates is large enough, the same pattern can be reproduced quite closely. The original manuscript uses ensembles of 100,000 replicates, which I have reproduced here.  A possible source of confusion is that the code defaults to only 100 replicates if the ensemble size is not explicitly verified -- which is an ensemble small enough to still see random deviations from the predicted results.   I probably set that lower default to facilitate faster testing.  Likewise, the MAXTIME parameter has to be extended from the default setting to allow the third initial condition set to run sufficiently long.

@@

Note that these tighter integration would also prevent the C libraries from running as purely stand-alone implementation.  In any event, meeting all the automated checks of a formal package is not an expectation of the compendium approach, which merely borrows some of the package tooling.

@@

In conclusion, I have been able to successfully replicate the original results.  As the original results were not performed with a fixed random seed, replicated results are statistically comparable but not bit-wise identical.  In a scientific sense this is perhaps more useful than replication with a fixed seed.  My compendium approach did not install entirely out of the box, as an autoconf configure script first required updating.  Ironically, the original version of the code published on a co-author's personal website and linked from the manuscript both still existed and did install out-of-the-box.  Reproducing the results of the paper required some careful reading of parameters listed in the paper and manually knowing what functions to call, as this paper did not include a complete "dynamic document" [@knitr; @knitr_book; @rmarkdown] approach discussed in @Marwick and also used in this replication paper.  I think this experience underscored the critical role dependency management has in frustrating reproducibility: this replication was relatively simple in part due to the low number of dependencies and the technical simplicity of the code.  The other take-away for me is how easy it is to leave out small but important steps needed to reproduce a figure when not using the dynamic document approach.  Despite having R scripts called "figure1.R" and "figure2.R", these scripts could not simply be run by themselves to create the figures.
